---
title: "Exam 2"
author: "Jillian Egland"
date: "2023-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r, message=FALSE}
library(ggplot2)
library(glmnet)
library(ISLR2)
library(MASS)
library(e1071)
```


```{r, results='hide'}
heart = read.table('https://hastie.su.domains/ElemStatLearn/datasets/SAheart.data', sep=",",head=T,row.names=1)
heart$famhis = as.factor(heart$famhist)
head(heart)
str(heart)

```


# Question 2: Coding 

## 2a 
probability of Y=1 when age=50: 
```{r} 
x=data.frame(age=50)
glm.fit = glm(chd~age, data=heart, family='binomial')
#summary(glm.fit)
#heart[heart$age==50,]
#head(glm.fit$fitted.values)

glm.prob = predict(glm.fit,heart[heart$age==50,],type='response') 
glm.prob = predict(glm.fit,x,type='response') 

prob_star = (glm.prob)[[1]]
prob_star





```

## 2b 
standard error of P(Y=1|age=50):  
```{r}
x=data.frame(age=50)
## bootstrap standard errors
n = dim(heart)[1]
B = 1000
prob = rep(0,B)
for(b in 1:B){
  index = sample(1:n,n,replace=TRUE)
  bootsample = heart[index,]
  glm.fit = glm(chd~age, data=bootsample, family='binomial')
  #glm.prob = predict(glm.fit,bootsample[bootsample$age==50,],type='response') 
  glm.prob = predict(glm.fit,x,type='response') 
  prob[b] = glm.prob
}

sqrt(sum((prob-mean(prob))^2)/(B-1))


```


## 2c 
confidence interval for P(Y=1|age=50):  
```{r}
##bootstrap confidence intervals
#p_star = above
se_prob_star = 0.026 
x=data.frame(age=50)

B = 100
m = 100
Fstar = rep(0,B)
prob_m = rep(0,m)

for(b in 1:B){
  index = sample(1:n,n,replace=TRUE)
  bootsample=heart[index,]
  #fit = lm(mpg~horsepower,data=bootsample)
  #beta0 =  coef(fit)[1]
  glm.fit = glm(chd~age, data=bootsample, family='binomial')
  glm.prob = predict(glm.fit,x,type='response') 
  prob = glm.prob
  
  for(i in 1:m){
    index2 = sample(index,n,replace=TRUE)
    bootsample2 = heart[index2,]
    #fit2 = lm(mpg~horsepower,data=bootsample2)
    #beta0_m[i] = coef(fit2)[1]
    glm.fit = glm(chd~age, data=bootsample2, family='binomial')
    #glm.prob = predict(glm.fit,bootsample2[bootsample2$age==50,],type='response') 
    glm.prob = predict(glm.fit,x,type='response') 
    prob_m[i] = glm.prob
    #prob_m[i] = (glm.prob)[[1]]
  }
  se_prob = sqrt(sum((prob_m-mean(prob_m))^2)/(m-1))

  Fstar[b] = (prob - prob_star)/se_prob
  #print(Fstar[b])
}


#95% confidence interval
#quantile(Fstar,c(0.025,0.975))
prob_star + quantile(Fstar,0.025)*se_prob_star
prob_star + quantile(Fstar,0.975)*se_prob_star



```


## 2d
Plot below shows that there seems to be a difference in the slopes of tobacco vs chd diagnosis depending on family history.  
```{r}
ggplot(heart, aes(x=tobacco, y=chd, colour = famhist)) + geom_point() + geom_smooth(method = "lm", fill = NA)

```

Misclassification rate for model without interaction:  
```{r}
n = dim(heart)[1]
train_index = sample(1:n,n/2,replace=F)
train = heart[train_index,]
test = heart[-train_index,] 


glm.fit = glm(chd ~ age  + famhist, data=train, family='binomial') ## specify your dataset
#summary(glm.fit)

glm.prob = predict(glm.fit,test,type='response') 
glm.pred = rep(0,nrow(test))
glm.pred[glm.prob >0.5] =1
table(glm.pred,test$chd)
mean(glm.pred != test$chd)

```
false negative rate: .605  


With interaction:  
```{r}
n = dim(heart)[1]
#train_index = sample(1:n,n/2,replace=F)
train = heart[train_index,]
test = heart[-train_index,] 


glm.fit.interact = glm(chd ~ age + tobacco + famhist + tobacco*famhist, data=heart,family='binomial') 
#summary(glm.fit)

glm.prob.interact = predict(glm.fit.interact,test,type='response') 
glm.pred.interact = rep(0,nrow(test))
glm.pred.interact[glm.prob.interact >0.5] =1
table(glm.pred.interact,test$chd)
mean(glm.pred.interact != test$chd)

```
false negative rate: .568  

It is worse to say someone doesn't have chd when they actual do have chd so we want to have a lower false negative rate. The 2 models have similar misclassification rates based on the training and test split, so I would look at the false negative rate. The model with the interaction has a lower false negative rate, so I would choose that model.  









